---
title: "NEON forecast challenge submission"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## First Forecast: Water Temp forecast using Single Layer Neural Network

Single layer neural network (mlp from tidy models). 
 - Trained model using historic data from NOAA 
 - Each predictor variable is multiplied by a weight representing importance 
 - Hypertuned model: 'penalty' and 'epochs' to run the training data 100 times 
 - Added process uncertainty (sigma) and driver uncertainty.

Using air temperature, relative humidity, and precipitation flux in a regression model to forecast for water temperature.

```{r load-packages, echo = F, warning=F, message=F}
## install.packages('remotes')
## install.packages('fpp3') # package for applying simple forecasting methods
## install.packages('tsibble') # package for dealing with time series data sets and tsibble objects
## install.packages('tidyverse') # collection of R packages for data manipulation, analysis, and visualisation
## install.packages('lubridate') # working with dates and times
## remotes::install_github('eco4cast/neon4cast') # package from NEON4cast challenge organisers to assist with forecast building and submission
install.packages("tidymodels")

# Load packages
library(tidyverse)
library(lubridate)
library(neon4cast)
library(tidymodels)

set.seed(215)

```

Read in target datasets

```{r get-targets, message=F}
#read in the targets data
targets <- read_csv('https://data.ecoforecast.org/neon4cast-targets/aquatics/aquatics-targets.csv.gz')

# read in the sites data
aquatic_sites <- read_csv("https://raw.githubusercontent.com/eco4cast/neon4cast-targets/main/NEON_Field_Site_Metadata_20220412.csv") |>
  dplyr::filter(aquatics == 1)

lake_sites <- aquatic_sites %>%
  filter(field_site_subtype == 'Lake')

# Filter the targets
targets <- targets %>%
  filter(site_id %in% lake_sites$field_site_id,
         variable == 'temperature')
```

Read in past NOAA data. In this model I am using air temperature, relative humidity, and precipitation flux to predict water temperature.

```{r get-NOAA-past, message = F}
# Past stacked weather
noaa_past_s3 <- neon4cast::noaa_stage3()

variables <- c("air_temperature", "relative_humidity", "precipitation_flux")

#Other variable names can be found at https://projects.ecoforecast.org/neon4cast-docs/Shared-Forecast-Drivers.html#stage-3

noaa_past <- noaa_past_s3  |> 
  dplyr::filter(site_id %in% lake_sites$field_site_id,
                datetime >= ymd('2017-01-01'),
                variable %in% variables) |> 
  dplyr::collect()

# aggregate the past to mean values
noaa_past_mean <- noaa_past |> 
  mutate(datetime = as_date(datetime)) |> 
  group_by(datetime, site_id, variable) |> 
  summarize(prediction = mean(prediction, na.rm = TRUE), .groups = "drop") |> 
  pivot_wider(names_from = variable, values_from = prediction) |> 
  # convert air temp to C
  mutate(air_temperature = air_temperature - 273.15)
```

```{r get-NOAA-future, message = F}
forecast_date <- Sys.Date() 
noaa_date <- forecast_date - days(1)

noaa_future_s3 <- neon4cast::noaa_stage2(start_date = as.character(noaa_date))
variables <- c("air_temperature", "relative_humidity", "precipitation_flux")

noaa_future <- noaa_future_s3 |> 
  dplyr::filter(datetime >= forecast_date,
                site_id %in% lake_sites$field_site_id,
                variable %in% variables) |> 
  collect()

noaa_future_daily <- noaa_future |> 
  mutate(datetime = as_date(datetime)) |> 
  # mean daily forecasts at each site per ensemble
  group_by(datetime, site_id, parameter, variable) |> 
  summarize(prediction = mean(prediction)) |>
  pivot_wider(names_from = variable, values_from = prediction) |>
  # convert to Celsius
  mutate(air_temperature = air_temperature - 273.15) |> 
  select(datetime, site_id, air_temperature, parameter, relative_humidity, precipitation_flux)

#Set up targets dataframe
targets_df <- targets |> 
  filter(variable == 'temperature') |>
  pivot_wider(names_from = 'variable', values_from = 'observation') |> 
  left_join(noaa_past_mean, 
            by = c("datetime","site_id")) |> 
  mutate(doy = yday(datetime))
```

```{r mlp-model-setup}
#split data by site, 80-20 split from targets_df
split <- initial_split(targets_df, prop = 0.80, strata = site_id)

train_data <- training(split)
test_data <- testing(split)

# define recipe
m_recipe <- train_data |> 
  recipe(temperature ~ . ) |> 
  step_rm(datetime) |>
  step_rm(site_id) |> 
  step_naomit(air_temperature, relative_humidity, precipitation_flux)

m_recipe

model <- mlp(engine = "nnet",
             #penalty = balance of model against training data. manually adjusted
             penalty = 0.01,
             #epochs = # of cycles through training data
             epochs = 100)|> 
  set_mode("regression")

workflow <- workflow() |> 
  add_model(model) |> 
  add_recipe(m_recipe)

fit <- workflow |> 
  fit(data = train_data)

predictions <- predict(fit, new_data = test_data)
pred_test <- bind_cols(test_data, predictions)

#Evaluate model using metrics table
multi_metric <- metric_set(rmse, rsq)

metric_table <- pred_test |> 
  multi_metric(truth = temperature, estimate = .pred) 

#View metrics table from model run
metric_table
```

```{r forecast-loop}
#residual for each ensemble member
residuals <-  pred_test$.pred - pred_test$temperature

#Process uncertainty
sigma <- sd(residuals, na.rm = TRUE)

targets_future <- noaa_future_daily |> 
  mutate(temperature = NA,
         doy = yday(datetime)) |> 
  filter(parameter == 1) |> 
  select(-parameter)

new_predictions <- predict(fit, new_data = targets_future)

targets_future <- noaa_future_daily |> 
  mutate(temperature = NA,
         doy = yday(datetime))

tidymodels_forecast <- data.frame()

for(i in unique(targets_future$parameter)){
  curr_ens <- targets_future |> 
    filter(parameter == i) |> 
    select(-parameter)
  
  new_predictions <- predict(fit, new_data = curr_ens)
  #adding process uncertainty to each ensemble members
  new_predictions$.pred = new_predictions$.pred + rnorm(n=length(new_predictions$.pred),mean=0,sd=sigma)
  curr_ens <- bind_cols(curr_ens, new_predictions) |> 
    mutate(parameter = i)
  tidymodels_forecast <- bind_rows(tidymodels_forecast, curr_ens)
}

```

# 

```{r}
tidymodels_forecasts_EFI <- tidymodels_forecast %>%
  rename(prediction = .pred) %>%
  mutate(variable = "temperature") |> 
  # For the EFI challenge we only want the forecast for future
  filter(datetime > Sys.Date()) %>%
  group_by(site_id, variable) %>%
  mutate(reference_datetime = min(datetime) - lubridate::days(1),
         family = "ensemble",
         model_id = "tidymodels_lm") %>%
  select(model_id, datetime, reference_datetime, site_id, family, parameter, variable, prediction)

tidymodels_forecasts_EFI |>
  filter(variable == "temperature") |>
  ggplot(aes(x = datetime, y = prediction, group = parameter)) +
  geom_line() + 
  facet_wrap(~site_id)
```

```{r make-standard}
# Make forecast fit the EFI standards
my_model_id <- 'mlp1_wtempforecast_LF'

forecast_df_EFI <- tidymodels_forecasts_EFI %>%
  filter(datetime > forecast_date) %>%
  mutate(model_id = my_model_id,
         reference_datetime = forecast_date,
         family = 'ensemble',
         parameter = as.character(parameter)) %>%
  select(datetime, reference_datetime, site_id, family, parameter, variable, prediction, model_id)

```

```{r write-forecast}
# Write the forecast to file
theme <- 'aquatics'
date <- forecast_df_EFI$reference_datetime[1]
forecast_name_1 <- paste0(forecast_df_EFI$model_id[1], ".csv")
forecast_file_1 <- paste(theme, date, forecast_name_1, sep = '-')
forecast_file_1


if (!dir.exists('Forecasts')) {
  dir.create('Forecasts')
}

write_csv(forecast_df_EFI, file.path('Forecasts',forecast_file_1))

```

```{r submit-forecast}

## # can uses the neon4cast::forecast_output_validator() to check the forecast is in the right format

# UNCOMMMENT THIS WHEN YOU ARE READY TO SUBMIT
neon4cast::submit(forecast_file = file.path('Forecasts', forecast_file_1),
                  ask = FALSE) # if ask = T (default), it will produce a pop-up box asking if you want to submit


```

```{r plot-forecast}
forecast_df_EFI |> 
  ggplot(aes(x=datetime, y=prediction, group = parameter)) +
  geom_line() +
  facet_wrap(~site_id) +
  labs(title = paste0('Forecast generated for ', forecast_df_EFI$variable[1], ' on ', forecast_df_EFI$reference_datetime[1]))
```
